\chapter{Conclusion and Future Work}
\label{ch:conclusion}

\section{Summary of the Thesis}
\label{sec:conclusion-summary}

This thesis studied two questions about large neural models: where a capability sits
inside a model, and how the training objective changes what the model does. We
worked on these across three studies, two in multilingual language models and one in
multi-label classification, and the studies pointed in a consistent direction.
Acting on a localized part of a model does not reliably change its behaviour, while
acting through the training objective does, though only where the training signal can
actually reach.

Chapter~\ref{ch:neurons} tested whether language-specific neurons can be used to
improve cross-lingual transfer. We identified these neurons in Llama~3.1 and Mistral
Nemo with two methods, LAPE and activation-probability thresholding, and tried both
test-time edits and neuron-targeted fine-tuning on XNLI and XQuAD. The interventions
did not improve performance on low-resource languages, and several of them degraded
it. We also found that the neurons are shared across languages, that setting their
activations to zero does not turn them off, and that they do no better than random
neurons under fine-tuning. The two identification methods place the neurons in
different parts of the model yet reach the same result, which suggests the negative
finding is about how language is represented rather than how the neurons are picked.

Chapter~\ref{ch:grpo-crosslingual} moved from a fixed model to training dynamics. We
fine-tuned Qwen2.5-7B-Instruct on the MGSM math benchmark with GRPO and tracked the
model across checkpoints with five measurements. GRPO changed accuracy very little
for any of the seven languages, and Telugu stayed far below the rest throughout. The
model reads out as English through the early and middle layers and surfaces the
target language only in the last few layers, but this structure was present from the
start and did not change over training, so GRPO did not cause or deepen cross-lingual
collapse. The one language that did not improve, Telugu, was also the one whose
representations training moved the most. We connected this to tokeniser coverage:
when a language has very few dedicated tokens, the reward signal is too coarse to
give a useful gradient, so training disturbs the representations without improving the
task.

Chapter~\ref{ch:fairpo} turned to designing an objective rather than analysing one.
We proposed FairPO, a framework for multi-label classification that splits the labels
into a privileged group to improve and a non-privileged group to maintain. The
privileged group is trained with a preference loss that prefers a true label over the
label most likely to be confused with it, with a BCE fallback for stability; the
non-privileged group is held near a reference model by a constrained loss; and the
two are balanced with a Group Robust Preference Optimization formulation. On MS-COCO
and NUS-WIDE, the CPO variant improved the rare privileged labels while leaving the
frequent labels almost unchanged, and the ablations showed that each part of the
framework contributes to this.

\section{Main Findings}
\label{sec:conclusion-findings}

Taken together, the three studies support a few points.

\begin{itemize}
	\item Editing a small, localized set of internal components is not a reliable way
	to change a model's cross-lingual behaviour. This held for neuron interventions
	in Chapter~\ref{ch:neurons}, and it is consistent with
	Chapter~\ref{ch:grpo-crosslingual}, where the English-leaning middle of the
	stack, the localized mechanism the collapse hypothesis points to, was not what
	GRPO acted on and not what decided the outcome.
	
	\item The training objective is the more effective place to act, but its effect
	is bounded by whether the training signal can reach the model at a fine enough
	level. GRPO improved a language only when the tokeniser gave it enough
	resolution, and FairPO improved the privileged labels because its objective acted
	directly on the pairs of labels that needed separating.
	
	\item A method that targets a chosen part of the label space, and protects the
	rest, can improve the targeted part without measurable harm to the rest, as
	FairPO showed. This is the constructive side of the same idea that the two
	analysis chapters reached from the negative side.
\end{itemize}

\section{Limitations}
\label{sec:conclusion-limitations}

The studies have limitations, which are stated in full in each chapter and collected
here. The neuron study looks only at the MLP layers and not at attention, uses two
models and two tasks, and relies on statistics from Wikipedia text for the test-time
interventions. The GRPO study uses one model family and one task, fine-tunes a
low-rank adapter on the attention projections only, and depends on a logit-lens
readout and a judge-based reward, both of which are imperfect; the tokeniser-coverage
counts are also approximate. The FairPO study uses a frequency-based label partition
on two image benchmarks, depends on a confusing set that changes during training, and
needs care in tuning the group-robust balancing. None of the three studies makes
claims beyond the models, languages, tasks, and datasets it uses.

\section{Future Work}
\label{sec:conclusion-future}

Several directions follow from this work.

\paragraph{Cross-lingual transfer beyond neuron edits.}
The neuron study ruled out one approach but did not solve the underlying problem of
weak low-resource performance. The finding that language is spread across the network
rather than held in a few neurons suggests looking at interventions that act on
larger parts of the model, or on the training data and objective, rather than on
single neurons.

\paragraph{Tokeniser coverage and low-resource reasoning.}
The GRPO study points to tokeniser coverage as the factor that decides whether
training can teach a language. A direct next step is to test this more carefully, for
example by changing the tokeniser or adding tokens for a low-coverage language and
checking whether GRPO can then improve it. This would tell us whether coverage is the
cause or only a correlate.

\paragraph{Broader testing of the GRPO findings.}
The cross-lingual study used one model and one task. The same measurements could be
run on other model families, other reasoning tasks, and more languages, to see
whether the flat per-layer pattern and the coverage effect hold more generally.

\paragraph{Extending FairPO.}
FairPO was tested on image multi-label classification with a frequency-based
partition. It could be applied to other sources of disparity, such as label
importance set by domain knowledge, and to other settings, including generation
tasks, where the privileged and non-privileged groups would be defined over
attributes rather than labels. The stability of the confusing set and the tuning of
the group-robust balancing are also worth studying further, along with any
theoretical analysis of when the minimax objective converges.

